---
title: 'Stan'
teaching: 60
exercises: 4
---


```{r, include=FALSE}
library(magrittr)
library(tidyverse)
library(grafify)
library(cowplot)
library(cmdstanr)

# Install CmdStan if it isn't available
if (is.null(cmdstanr::cmdstan_version(error_on_NA = FALSE))) {
  cmdstanr::install_cmdstan(cores = 2)
}

cmdstan_setup <- function() {

  home_dir <- normalizePath("~", winslash = "/")
  work_dir <- file.path(home_dir, "cmdstan_work")
  temp_dir <- file.path(home_dir, "cmdstan_temp")

  dir.create(work_dir, FALSE, TRUE)
  dir.create(temp_dir, FALSE, TRUE)

  Sys.setenv(
    TMPDIR = temp_dir,
    TEMP   = temp_dir,
    TMP    = temp_dir
  )

  options(
    cmdstanr_write_stan_file_dir = work_dir,
    cmdstanr_write_exe_dir = work_dir
  )
}

cmdstan_setup()

cmdstanr::register_knitr_engine(override = TRUE)

theme_set(theme_bw(15))
posterior_color <- "#56B4E9"

```


::::::::::::::::::::: instructor
The goal of this episode is to teach the basic Stan workflow:

  1. Implement a model in a Stan program; 
  2. Feed in data and generate posterior samples; 
  3. Extract summaries or samples in R.

### What to emphasize

  - Stan generates the posterior samples, with cmdstanr providing the interface from R. Subsequent analysis is done in R.
    - MCMC is presented in a later episode.
  - The basic program blocks answer the following questions:
    - data: What do we observe?
    - parameters: What do we want to infer?
    - model: How are data and parameters related?
  - Parameter constraints can be important, and incorrectly specified constraints can prevent the program from working correctly.
  - The generated quantities block is where posterior predictions (and other derived quantities) are generated.
  - Posterior predictions naturally incorporate both parameter uncertainty and observation-level variability.
  - Simulation is an important model-development tool, and failure to reasonably recover the generating parameters can signal issues.

### Tips

  - Have learners install and test CmdStan before the lesson! Compiler problems are common and can consume substantial teaching time.
  - There is no need to spend much time on MCMC mechanics, diagnostics, Jacobians, or code optimization; these can be covered later.

::::::::::::::::::::::::::::::::


:::::::::::::::::::::::::::::::::::::: questions 

- How can posterior samples be generated using Stan?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

Learn how to

- implement statistical models in Stan; 
- generate posterior samples with Stan; 
- extract and process samples generated with Stan.

::::::::::::::::::::::::::::::::::::::::::::::::

Stan is a probabilistic programming language that can be used to specify probabilistic models and to generate samples from posterior distributions. 

The standard workflow when using Stan is to first write the statistical model in a separate text file, and then to call Stan from R using `cmdstanr`, which interfaces with the underlying CmdStan engine to perform the sampling. Instead of having to write explicit mathematical formulas the model is specified using built-in functions and sampling statements similar to written text. The sampling process is performed with a Markov Chain Monte Carlo (MCMC) algorithm, a topic of a later episode. For now, however, our focus is on understanding how to execute it using Stan. 

To get started, follow the instructions provided at **[https://mc-stan.org/cmdstanr/](https://mc-stan.org/cmdstanr/)** to install `cmdstanr` and the underlying CmdStan backend on your local computer.

::::::::::::::::::::::::::::::::::: callout

Several R packages have been built that simplify Stan usage. For example, brms allows specifying models via R's customary formula syntax, while bayesplot provides a library of plotting functions. In this lesson, however, we will first learn using Stan from the bottom-up, by writing Stan programs, extracting the posterior samples and performing subsequent analyses ourselves. Later, in episode 7, we'll introduce the usage of some of these additional packages. 

:::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::: callout

With Stan, you can fit models that have continuous parameters, but sampling from models with discrete parameters (e.g. clustering models) is not directly supported. In some cases, we can work around this by marginalizing out the discrete parameters, which makes the model fit possible in Stan. However, such workarounds are not always straightforward.

:::::::::::::::::::::::::::::::::::::::::::


## Basic program structure

A Stan program is organized into several blocks that collectively define the model. Typically, a Stan program includes at least the following blocks:

1. Data: This block is used to declare the input data. It specifies the types and dimensions of the data variables incorporated into the model.

2. Parameters: In this block, the model parameters are declared. 

3. Model: The likelihood and prior distributions are included here through sampling statements. 

For best practices, it is recommended to specify Stan programs in separate text files with a .stan extension, which can then be read, compiled, and sampled from R using `cmdstanr`.


## Example 1: Binomial model
  
The following Stan program specifies the Beta-binomial model, and consists of data, parameters, and model blocks. 

The data variables are the total sample size $N$ and $x$, the outcome of a binary variable (coin flip, handedness etc.). The declared data type is `int` for integer, and the variables have a lower bound 1 and 0 for $N$ and $x$, respectively. Notice that each line ends with a semicolon.

In the parameters block we declare $\theta$, the probability for a success. Since this parameter is a probability, it is a real number restricted between 0 and 1.

In the model block, the likelihood is specified with the sampling statement `x ~ binomial(N, theta)`. This line includes the binomial distribution $\text{Binomial}(x | N, theta)$ in the target distribution. The prior is set similarly, and omitting the prior implies a uniform prior. Comments can be included after two forward slashes. 
  

```{stan output.var="binomial_model"}
  data{
    int<lower=1> N; 
    int<lower=0> x; 
  }
  
  parameters{
    real<lower=0, upper=1> theta;
  }
  
  model{
    // Likelihood
    x ~ binomial(N, theta);
    
    // Uniform prior
  }
```

When the Stan program has been saved we need to compile it. In R, this is done by running the following line, where `"binomial_model.stan"` is the path of the program file. 

```{r, eval=FALSE}
binomial_model <- cmdstan_model("binomial_model.stan")
```

Once the program has been compiled, it can be used to generate the posterior samples by calling the `$sample()` method on the model object. The data need to be supplied as a list.

```{r, message=FALSE}
binom_data <- list(N = 50, x = 7)

binom_samples <- binomial_model$sample(data = binom_data,
                          seed = 135)
```


With the default settings, CmdStan (via `cmdstanr`) runs 4 MCMC chains. Each chain runs for 2000 total iterations (more about this in the next episode). During the run, CmdStan provides real-time progress information, aiding in estimating the running time, particularly for complex models or extensive datasets. In this case, the sampling took only a fraction of a second. 

Running `binom_samples$summary()`, a summary for the model parameter $\theta$ is printed, facilitating a quick review of the results.

```{r}
binom_samples$summary()
```


This summary is returned as a data frame, allowing easy manipulation using, for example, tidyverse tools. Often, however, it is necessary to process the individual samples. These can be extracted as a matrix using the $draws() from `cmdstanr` method:

```{r}
theta_samples <- binom_samples$draws("theta", format = "matrix")
```

Now we can use the methods presented in the previous Episode to compute posterior summaries, credible intervals and to generate figures. 

:::::::::::::::::::::::::::::::::::: challenge

Compute the 95% credible intervals from the samples drawn with Stan. What is the posterior probability that $\theta \in (0.05, 0.15)$? Plot a histogram of the posterior samples. 


::::::::::::::::::::: solution

```{r}

CI95 <- quantile(theta_samples, probs = c(0.025, 0.975))
theta_between_0.05_0.15 <- mean(theta_samples>0.05 & theta_samples<0.15)

plot_data <- data.frame(theta = as.vector(theta_samples))

p <- ggplot(data = plot_data) +
  geom_histogram(aes(x = theta), bins = 30) +
  coord_cartesian(xlim = c(0, 1))


print(p)
```


::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::::::::::::



:::::::::::::::::::::::::::::::::::: challenge

Try modifying the Stan program so that you add a $\text{Beta}(\alpha, \beta)$ prior for $\theta$.

Can you modify the Stan program further so that you can set the hyperparameters $\alpha, \beta$ as part of the data? What is the benefit of using this approach?


::::::::::::::::::::: solution

Modifying the data block so that it declares the hyperparameters as data (e.g. `real<lower=0> alpha;` and `real<lower=0> beta;`) allows the prior parameters to be specified from R rather than hard-coded in the Stan program. This allows changing prior parameters without having to modify and recompile the program.

::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::

## Additional Stan blocks

In addition to the data, parameters, and model blocks there are additional blocks that can be included in the program. 

1. Functions: For user-defined functions. This block must be the first in the Stan program. It allows users to define custom functions.

2. Transformed data: This block is used for transformations of the data variables. It is often employed to preprocess or modify the input data before it is used in the main model. Common tasks include standardization, scaling, or other data adjustments.

3. Transformed parameters: In this block, transformations of the parameters are defined. If transformed parameters are used on the left-hand side of sampling statements in the model block, the Jacobian adjustment for the posterior density needs to be included in the model block as well ([see](https://mc-stan.org/docs/stan-users-guide/reparameterization.html#changes-of-variables)). 

4. Generated quantities: This block is used to define quantities based on both data and model parameters. These quantities are not part of the model but are useful for post-processing. Each quantity is evaluated for every posterior sample, producing a corresponding posterior distribution.

We will make use of these additional structures in subsequent illustrations. 

## Example 2: Normal model

Next, let's implement the normal model in Stan. First we'll generate some data $X$ from a normal model with unknown mean and standard deviation parameters $\mu$ and $\sigma$:

```{r}
# Sample size
N <- 99

# Generate data with unknown parameters
unknown_sigma <- runif(1, 0, 10)
unknown_mu <- runif(1, -5, 5)

X <- rnorm(n = N,
           mean = unknown_mu,
           sd = unknown_sigma) 

normal_data <- list(N = N, X = X)
```

::::::::::::::::: callout
A common and useful step in model development is to simulate data from the model using known parameter values, fit the model to the simulated data, and compare the estimated parameters with the values used to generate the data. If the estimates do not align with the true parameters, this may indicate problems with the model specification, simulation procedure, identifiability, or fitting algorithm.
:::::::::::::::::::::::::


The Stan program for the normal model is specified in the next code chunk. It introduces a new data type (vector) and leverages vectorization in the likelihood statement which speeds up the sampling. In the end of the program, a generated quantities block is included which generates new data (X_tilde) from the posterior predictive distribution. As mentioned above, each X_tilde sample is generated by drawing a random realization from the normal distribution for each posterior sample $(\mu, \sigma)$.



```{stan output.var="normal_model"}
data {
  int<lower=0> N;
  vector[N] X;
}
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  // Vectorized likelihood
  // Equivalent to for(i in 1:N) X[i] ~ normal(mu, sigma);
  X ~ normal(mu, sigma);
  
  // Priors
  mu ~ normal(0, 1);
  sigma ~ gamma(2, 1);
}
generated quantities {
  real X_tilde;
  X_tilde = normal_rng(mu, sigma);
}
```

Let's fit the model to the data using `cmdstanr`: 

```{r, message=FALSE, results='hold'}
normal_samples <- normal_model$sample(data = normal_data)
```

Next, we'll extract posterior samples and generate a plot for the joint, and marginal posteriors. The true unknown parameter values are included in the plots in red. 

```{r}
# Extract parameter samples
par_samples <- normal_samples$draws(c("mu", "sigma"), format = "df")

# Full posterior
p_posterior <- ggplot(data = par_samples) + 
  geom_point(aes(x = mu, y = sigma)) +
  annotate("point", x = unknown_mu, y = unknown_sigma, 
           color = "red", size = 5)

# Marginal posteriors
p_marginals <- ggplot(data = par_samples %>% 
                         as.data.frame() %>% 
                         tidyr::pivot_longer(cols = c(mu, sigma))) + 
  geom_histogram(aes(x = value), bins = 40) + 
  geom_vline(data = data.frame(
    name = c("mu", "sigma"), 
    value = c(unknown_mu, unknown_sigma)
  ), 
  aes(xintercept = value), color = "red", linewidth = 1) +
  facet_wrap(~name, scales = "free")


p <- cowplot::plot_grid(p_posterior, p_marginals, ncol = 1)

print(p)
```

Let's also plot the posterior predictive distribution samples histogram and compare it to that of the data. 

```{r, fig.height = 4, fig.width = 5}
PPD <- normal_samples$draws("X_tilde", format = "df")$X_tilde
PPD <- data.frame(X_tilde = PPD)

obs_data <- data.frame(X = X)

p_PPD <- ggplot() + 
  geom_histogram(data = PPD, 
                 aes(x = X_tilde, y = after_stat(density)), 
                 bins = 30, fill = posterior_color) +
  geom_histogram(data = obs_data, 
                 aes(x = X, y = after_stat(density)), 
                 bins = 30, alpha = 0.5)

print(p_PPD)
```



## Example 3: Linear regression

:::::::::::::::::::::::::::::::::::: challenge

Write a Stan program for linear regression with one dependent variable. 

Generate data from a linear model and use the Stan program to estimate the intercept $\alpha$, slope $\beta$, and deviation term $\sigma$.

::::::::::::::::::::: solution

```{stan output.var="linear_model"}
data {
  int<lower=0> N; // Sample size
  vector[N] x; // x-values
  vector[N] y; // y-values
}
parameters {
  real alpha; // intercept
  real beta;  // slope
  real<lower=0> sigma; // noise
}

model {
  // Likelihood
  y ~ normal(alpha + beta * x, sigma);
  
  // Priors
  alpha ~ normal(0, 1);
  beta ~ normal(0, 1);
  sigma ~ inv_gamma(1, 1);
}

```


::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::


:::::::::::::::::::::::::::::::::::: challenge

Modify the program for linear regression with $M$ predictor variables. 

::::::::::::::::::::: solution

```{stan output.var="multiple_lm"}

data {
  int<lower=0> N; // Sample size
  int<lower=0> M; // Number of features
  matrix[N, M] x; // x-values
  vector[N] y; // y-values
}
parameters {
  real alpha; // intercept
  vector[M] beta;  // slopes
  real<lower=0> sigma; // noise
}

model {
  
  // Likelihood
  y ~ normal(alpha + x * beta, sigma);
  
  // Priors
  alpha ~ normal(0, 1);
  beta ~ normal(0, 1);
  sigma ~ inv_gamma(1, 1);
}

```


::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::





::::::::::::::::::::::::::::::::::::: keypoints 

- Stan is a tool for efficient posterior distribution sample generation. 
- A Stan program is specified in a separate `.stan` file consisting of code blocks, and can be compiled and sampled from R using `cmdstanr`.
- Subsequent analysis of the posterior samples is performed outside Stan.

After this episode, you should be able to use Stan to fit probabilistic models by:

- Writing Stan model code, including:
  - data types and declarations,
  - program blocks,
  - sampling statements, and
  - data, parameters, priors, the model, and generated quantities.
- Providing data to the model.
- Compiling and fitting the model.
- Accessing the model output and posterior samples.
- Manipulating posterior samples.
- Visualizing the results.

::::::::::::::::::::::::::::::::::::::::::::::::



## Resources

- Official release paper: [https://www.jstatsoft.org/article/view/v076i01](https://www.jstatsoft.org/article/view/v076i01)
- Stan User's Guide: [https://mc-stan.org/docs/stan-users-guide/](https://mc-stan.org/docs/stan-users-guide/)
- Stan Reference Manual: [https://mc-stan.org/docs/reference-manual/](https://mc-stan.org/docs/reference-manual/)
- Stan Functions Reference: [https://mc-stan.org/docs/functions-reference/](https://mc-stan.org/docs/functions-reference/)
- Stan Discourse Forum: [https://discourse.mc-stan.org](https://discourse.mc-stan.org)
- Stan Case Studies: [https://mc-stan.org/users/documentation/case-studies](https://mc-stan.org/users/documentation/case-studies)

## Reading

- BDA3: Ch. 12.6, Appendix C
- Bayes Rules!: Ch. 6.2 
